Summary of Survey Analysis NDNS carb50, 7 time slots

### 2018-07-31 paper/Chapters/Chapter2.tex
LCGA removed

To produce a sequence of discrete responses about the carbohydrate intake we are interested, we calculated the energy consumption per hour for each participant. The percentage of energy that contributed by carbohydrate within each hour were calculated for every every participant, each single day. Since we planed to apply latent class analysis (LCA) in the current study, in which the observed indicators for latent classes must be categorical, we then dichotomised the responses according to the carbohydrate contribution to the energy intake at cut-off value of 50\%, i.e. if within an hour there is any energy intake occured, carbohydrate consumption was categorised into whether it's energy contribution was lower or higher/equal to 50\% of total energy intake within that hour. Consequently, for each day of the recording, there were 24 data points generated by the dairy, each data point included one of the following responses:

\begin{itemize}
	\item Not eating any food; 
	\item Eating, and carbohydrate contributed less than 50\% of the total energy intake;
	\item Eating, and carbohydrate contributed higher or equal to 50\% of the total energy intake.
\end{itemize}
\vspace{-0.5cm}

@@

Overall, for years 1 to 8 combined, a sample of 39,300 address was selected from 1,438 PSUs. The sampling selection process was:

@@

Equation \ref{LCA} indicates that responses for an observation to the measuring variables are independent of one another given its membership in latent class $t$. However, in the NDNS RP data set, the assumption of independent observations is violated. Each individual completed their dietary diary for four consecutive days, their diary recordings were later converted into 4 sequences of categorical responses reflecting the type of carbohydrate consumption at each hour of the day. The 4 observed sequences (observations) are nested in the participants and therefore are not independent. This nested data structure requires multilevel techniques. 
\vspace{-0.3cm}

@@

Multilevel latent class analysis accounts for the nested structure of the data by allowing latent class intercepts to vary across level 2 units and thereby examining if and how level 2 units influence the level 1 latent classes. These random intercepts allow the probability of
membership in a particular level 1 (observations) latent class to vary across level 2 units (e.g., here in the current context are the individuals). Essentially this allows the probability that an observation will belong to a
particular level 1 latent class to vary across Level 2 units (individuals). \vspace{-0.5cm}

@@

Proposed by Vermunt \parencite{Vermunt, vermunt2008latent} and Asparouhov and Muth\'en \parencite{muthen2009multilevel},  a traditional, parametric approach can be applied using a logistic regression model. In an unconditional logistic regression model, the probability of the outcome (i.e. being in latent class $k$) is constant within the 4-day survey for each individual (level 2). Therefore, say when we are fitting a model with $k (k = 1, \cdots, K)$ latent classes in level 1, then in each individual (level 2), there is a probability of being in latent class $k$. A random effect model consider the individual (level 2) to be drawn from a population of adults in the UK, and the probability of the outcome (i.e. being in latent class $k$) across individuals is considered to be a random variable \parencite{snijders2011multilevel}. The 2-level random intercept effect regression model can be expressed as:\vspace{-0.4cm}

\begin{equation}
\begin{aligned}
\text{logit}[P(C_{ij} = t)] & = \beta_{0j} & \textbf{(level 1)}  \\
\beta_{0j} & = \gamma_0 + \gamma_1 w_j + u_{0j} & \textbf{(level 2)} \\ 
\Rightarrow P(C_{ij} = t) & = \frac{\exp{(\gamma_0 + \gamma_1 w_j + u_{0j})}}{1 + \exp{(\gamma_0 + \gamma_1 w_j + u_{0j})}} \\
\end{aligned}
\label{randomLCA}
\end{equation}
\vspace{-0.3cm}

@@

\begin{itemize}
	\item $P(C_{ij} = t)$ as the probability that the randomly selected $i$th observation of $j$th individual is belonging to latent class $t$;
	\item $u_{0j}$ as the random intercept for $j$th individual; 
	\item the random intercept are assumed be normally distributed (i.e. $u_{0j} \sim N(0, \sigma_{u_0}^2)$), the magnitude of the $u_{0j}$ variance ($\sigma_{u_0}^2$) indicates the influence of the individuals (level 2);
	\item $w_j$ is the predictor for individual (level 2), such as age, and/or sex.
\end{itemize}

Same as in the typical LCA models, the latent class variable in a MLCA is defined by multiple observed indicators (here is defined by the responses of eating carbohydrate within each hour, over 24 hours and during 4 consecutive days of survey period). Considering that the latent class indicators are indicator variables ($U_{ijk}$), the MLCA model can be written as follows:\vspace{-0.8cm}

\begin{equation}
%\begin{aligned}
P(U_{ij1} = s_1, U_{ij2} = s_2, \cdots, U_{ijk} = s_{K}) = \sum_{t=1}^{T}P(C_{ij}=t)\prod_{k=1}^{K}P(U_{ijk} = s_k | C_{ij} = t)
%\end{aligned}
\label{MLCA}
\end{equation}
\vspace{-0.8cm}

@@

\begin{itemize}
	\item $ U_{ijk} $ represents the response of eating \textcolor{red}{high/low} carbohydrate on $i$th day of the survey ($i \in (1,2,3,4)$) in subject $j$ (level 2) at the $k$th hour of the day ($k \in (1, 2, 3, \cdots, 24)$);
	\item $C_{ij}$ denotes the latent class membership for subject $j$ on $i$th day of the survey;
	\item A specific latent class is referred to as $t$, and the total number of level 1 latent classes is denoted by $T$;
	\item $P(U_{ijk} = s_k|C_{ij} = t)$ is the probability of a specific response pattern, conditional on membership in latent class $t$.
\end{itemize}

The $P(C_{ij} = t)$ in equation \ref{MLCA} is what we have already defined in equation \ref{randomLCA}.

@@

The number of classes in either level 1 were determined by 1) the evaluation of model fit indices, including the Akaike information criterion (AIC), Bayesian information criterion (BIC), adjusted Bayesian information criterion (aBIC) where smaller values indicate better, and entropy which is a statistic that summarizes latent class probabilities where values near 1 indicate better latent class separation; 2) the Lo-Mendell-Rubin Likelihood Ratio Test (LMR-LRT) \parencite{lo2001testing, nylund2007deciding} which compare $q$ vs. $q-1$ class models, where $q$ is the number of latent classes and 3) pattern interpretability. In the step of performing multilevel LCA, where neither LMR-LRT nor BS-LRT were available, same rules of model fit indices and pattern interpretability were used to determine the optimal combination of latent classes in level 1 and level 2.

\section{Latent Class Growth Analysis (LCGA)}\vspace{-0.3cm}

>>>>

Time across a typical survey day was divided into 7 time slots in the dietary diary of NDNS RP: 6 am to 9 am, 9 am to 12 noon, 12 noon to 2 pm, 2 pm to 5 pm, 5 pm to 8 pm, 8 pm to 10 pm, and 10 pm to 6 am. To produce a sequence of discrete responses regarding the carbohydrate intake we are interested, the energy consumption within each time slot over the four days of survey for each participant were calculated. The percentages of energy that contributed by carbohydrate within each time slot were then estimated. Since we planed to apply latent class analysis (LCA) in the current study, in which the observed indicators for latent classes must be categorical, we then dichotomised the responses according to the carbohydrate contribution to the energy intake at cut-off value of 50\%, i.e. if within a time slot there is any energy intake occured, carbohydrate consumption was categorised into whether it's energy contribution was lower or higher/equal to 50\% of total energy intake within that time slot. Consequently, for each day of the recording, there were 7 data points generated by the diary, each data point included one of the following responses:

\begin{itemize}
	\item Not eating any food (Energy intake = 0 kJ); 
	\item Eating, and carbohydrate contributed less than 50\% of the total energy intake;
	\item Eating, and carbohydrate contributed higher or equal to 50\% of the total energy intake.
\end{itemize}
\vspace{-0.5cm}

@@

Overall, for years 1 to 8 combined, a sample of 39,300 addresses was selected from 1,438 PSUs. The sampling selection process was:

@@

Equation \ref{LCA} indicates that responses for an observation to the measuring variables are independent of one another given its membership in latent class $t$. However, in the NDNS RP data set, the assumption of independent observations is violated. Each individual completed their dietary diary for four consecutive days, their diary recordings were later converted into four sequences of categorical responses reflecting the type of carbohydrate consumption at each time slot of the day. The four observed sequences (observed days) are nested in the participants and therefore are not independent. This nested data structure requires multilevel techniques. 
\vspace{-0.3cm}

@@

Multilevel latent class analysis accounts for the nested structure of the data by allowing latent class intercepts to vary across level 2 units and thereby examining if and how level 2 units influence the level 1 latent classes. These random intercepts allow the probability of membership in a particular level 1 (observation days) latent class to vary across level 2 units (e.g., here in the current context are the individuals). Essentially this allows the probability that an observation day will belong to a
particular day-level latent class to vary across individual levels.\vspace{-0.5cm}

@@

Proposed by Vermunt \parencite{Vermunt, vermunt2008latent} and Asparouhov and Muth\'en \parencite{muthen2009multilevel},  a traditional, parametric approach can be applied using a logistic regression model. For example, let's assume that there are two types of observation days in the dietary survey--high and low carbohydrate eating days. In an unconditional logistic regression model, the probability of the outcome (i.e. an observed high carbohydrate eating day vs. a low carbohydrate eating day) is constant within individual-level which means for each person throughout his/her survey there is some probability of following a high carbohydrate eating day. A random effect model considers the individuals (level 2) to be drawn from the adult population in the UK, and the probability of the outcome (i.e. high carbohydrate eating days) across individuals is considered to be a random variable \parencite{snijders2011multilevel}.

Thus, for a binary outcome $C_{ij} = 0, 1$ (low $=0$ or high $=1$ carbohydrate eating days), where $i$ denotes the observation days $(i = 1, 2, 3, 4)$, and $j$ denotes the individual $(j = 1, 2, \cdots, 6155)$. The 2-level random intercept logistic regression model can be expressed as:\vspace{-0.4cm}

\begin{equation}
\begin{aligned}
\text{logit}[P(C_{ij} = 1)] & = \beta_{0j} + \beta_{1}x_{ij} \;\;\;\;\;\;\;\;\; \textbf{(day level)}  \\
\beta_{0j} & = \gamma_0 + \gamma_1 w_j + u_{0j} \; \textbf{(individual level)} \\ 
\Rightarrow P(C_{ij} = 1) & = \frac{\exp{(\gamma_0 + \beta_{1}x_{ij} + \gamma_1 w_j + u_{0j})}}{1 + \exp{(\gamma_0 + \beta_{1}x_{ij} + \gamma_1 w_j + u_{0j})}} \\
\end{aligned}
\label{randomLCA}
\end{equation}
\vspace{-0.3cm}

@@

\begin{itemize}
	\item $P(C_{ij} = 1)$ as the probability that the randomly selected $i$th observation day of $j$th individual is a high carbohydrate eating day;
	\item $\beta_{0j}$ as the random intercept, for outcome $C_{ij} = 1$; 
	\item the random deviation of the individuals $u_{0j}$ are assumed be normally distributed (i.e. $u_{0j} \sim N(0, \sigma_{u_0}^2)$), the magnitude of the $u_{0j}$ variance ($\sigma_{u_0}^2$) indicates the influence of the individuals (level 2);
	\item $x_{ij}, w_j$ is the predictors for day level (weekdays or weekends) and individual level, such as age, and sex.
\end{itemize}

Same framework can be used to consider random effects in an LCA model, but instead of saying that $C_{ij}$ is either low or high carbohydrate eating days as if we already know, it is now replaced by a latent variable $G_{ij}$ which indicates the typologies of carbohydrate eating patterns. Then we can use the day-level data to assess the log-odds of belonging to $k$th type of carbohydrate eating pattern on a specific day of survey, and we allow the log-odds to vary across individuals. Therefore, for some persons the log-odds of having a $k$th type of carbohydrate eating pattern during the survey can be high, but for the other persons, the log-odds of following the $k$th type of carbohydrate eating pattern can be low.

If the day-level LCA model (carbohydrate eating pattern typologies) is best defined by $T (T \geqslant 2)$ latent classes, then $T-1$ random intercept will be specified by a two-level multinomial logistic regression model. Similar to the typical LCA models, the latent class variable in a MLCA is defined by multiple observed indicators (here is defined by the responses of eating carbohydrate within each time slots, throughout 4 consecutive days of survey period). Considering the latent class indicators are indicator variables ($U_{ijk}$), the MLCA model can be written as follows:\vspace{-0.8cm}

\begin{equation}
%\begin{aligned}
P(U_{ij1} = s_1, U_{ij2} = s_2, \cdots, U_{ijk} = s_{K}) = \sum_{t=1}^{T}P(G_{ij}=t)\prod_{k=1}^{K}P(U_{ijk} = s_k | G_{ij} = t)
%\end{aligned}
\label{MLCA}
\end{equation}
\vspace{-0.8cm}

@@

\begin{itemize}
	\item $ U_{ijk} $ represents the response of eating carbohydrate (one of the following: not eating any food, $< 50\%$ of the energy, or $\geqslant 50\%$ of the energy) on $i$th day of the survey ($i \in (1,2,3,4)$) in $j$th individual at the $k$th time slot of the day ($k \in (1, 2, 3, \cdots, 7)$);
	\item $G_{ij}$ denotes the latent class membership for $j$th individuals on $i$th day of the survey, the total number of day-level latent class is $T$;
%	\item A specific latent class is referred to as $t$, and the total number of level 1 latent classes is denoted by $T$;
	\item $P(U_{ijk} = s_k|C_{ij} = t)$ is the probability of a specific response pattern, conditional on membership in latent class $t$.
\end{itemize}

The $P(G_{ij} = t)$ in equation \ref{MLCA} is what we have already defined in equation \ref{randomLCA}: \vspace{-0.3cm}

\begin{equation}
P(G_{ij} = t) = \frac{\exp{(\gamma_0 + \beta_{1}x_{ij} + \gamma_1 w_j + u_{0j})}}{1 + \exp{(\gamma_0 + \beta_{1}x_{ij} + \gamma_1 w_j + u_{0j})}} 
\end{equation}

@@

The number of classes in either level 1 were determined by 1) the evaluation of model fit indices, including the Akaike information criterion (AIC), Bayesian information criterion (BIC), adjusted Bayesian information criterion (aBIC) where smaller values indicate better, and entropy which is a statistic that summarizes latent class probabilities where values near 1 indicate better latent class separation; 2) the Lo-Mendell-Rubin Likelihood Ratio Test (LMR-LRT) \parencite{lo2001testing, nylund2007deciding} which compare $q$ vs. $q-1$ class models, where $q$ is the number of latent classes and 3) pattern interpretability. In the step of performing multilevel LCA, where neither LMR-LRT nor BS-LRT were available, same rules of model fit indices and pattern interpretability were used to determine the optimal combination of latent classes in level 1 and level 2. \vspace{-0.3cm}

%\section{Latent Class Growth Analysis (LCGA)}\vspace{-0.3cm}
%
%%LCGA is a statistical approach that models patterns in longitudinal data by classifying individuals into unobserved latent classes with more homogeneous patterns. LCGA is a special type of Growth Mixture Models (GMM), in which the variance of latent slope and intercept are fixed to zero within class, and allowed to vary only across classes. LCGA assumes that all individual growth trajectories within classes are homogeneous. 
%
%
%LCGA is a semi‐parametric technique used to identify distinct subgroups of individuals following a similar pattern of change over time on a given variable. The analysis identifies distinct subgroups of individuals following a distinct pattern of change over time.  LCGA is a special type of Growth Mixture Models (GMM) \parencite{jung2008introduction}, in which the variance of latent slope and intercept are fixed to zero within class, and allowed to vary only across classes. LCGA assumes that all individual growth trajectories within classes are homogeneous. The diagramme of the LCGA model in the current setting is illustrated in \textbf{Figure \ref{fig:LCGAdiagram}}:
%
%\begin{figure}[H]
%	%\vspace*{13cm}
%	\centering
%	\includegraphics[width=10cm]{Figures/LCGAdiagramme.jpeg}
%	\decoRule
%	\caption[Diagramme for LCGA model]{Diagram for LCGA model.\\ \scriptsize{\textbf{Abbreviations:} g, latent class variable; i, intercept; l, linear term coefficient; q, quadratic term coefficient; c, cubic term coefficient; u1, u2, ... u96, repeated categorical outcomes of eating carbohydrates at each hour/time slots across four days of survey.}}
%	\label{fig:LCGAdiagram}
%\end{figure}\vspace{-0.6cm}
%
%In the NDNS RP data context, we considered a latent categorical variable $g_i$ representing the unobserved subpopulation membership (latent class variable) for participant $i, (g_{i} = 1,2,\cdots K)$. Then a LCGA model predicts the probability of the carbohydrate eating response $u_{ti}$ at time $t$ in latent class $K$ can be written as: \vspace{-0.3cm}
%
%\begin{equation}
%\begin{aligned}
%P(u_{ti} = 0/1/2 |g_i = K) &  = \frac{1}{1 + e^{-\text{logit}(u_{itK})}} \\
%\text{logit}(u_{itK}) & = \beta_{0K} + \beta_{1K}t_{i} + \beta_{2K}t^2_i + \beta_{3K}t^3_i + \epsilon_{i} \\ 
%\epsilon_{i} \sim N(0, \sigma^2) & \;i = 1,2,...6155
%\end{aligned}
%\end{equation}\vspace{-0.6cm}
%
%Where, 
%
%\begin{itemize}
%	\item $u_{ti}=\left\{ \begin{array}{ll}  0 & \text{if particiant } i \text{ was not eating any food at time } t\\ 
%	 1 & \text{if particiant } i \text{ was eating food and contribution of energy} \\ 
%	   &  \text{from carbohydrate was lower than 50\% at time } t \\ 
%	 2 & \text{if participant } i \text{ was eating food and contribution of energy} \\ 
%	 &  \text{from carbohydrate was higher or equal to 50\% at time } t \end{array} \right.$
%	\item $\beta_{0K}, \beta_{1K}, \beta_{2K}, \beta_{3K}$  correspond to $i, l, q, c$ in \textbf{Figure \ref{fig:LCGAdiagram}}, i.e. the coefficients for intercept, linear, quadratic, and cubic term of time.
%\end{itemize}
%
%An LCGA model can be modelled using either a linear, quadratic, or cubic trend over time. However, in our case, it may not be appropriate to assume the probabilities of responses to eating carbohydrates on the logit scale to change on a linear, or quadratic trend over four days, we chose to fit a more flex time function-a cubic form of time. The assumption in a LCGA model is that the data trajectory (the curve of eating carbohydrate pattern) for the participant consisting of the repeated measurements over time $t$, are independent given the group, $g_i$.

### 2018-07-31 results/table1slots.Rmd
LCGA removed

Energy14 <- food14 %>% 
  select(ID, Country, SurveyYear, EnergykJ, Carbohydrateg, CHOpctotE)
Energy14$Country[Energy14$Country == "Northern Ireland"] <- "NI"

Energy56 <- food56 %>% 
  select(ID, Country, Surveyyear, EnergykJ, Carbohydrateg, CHOpctotE) %>% 
  rename(SurveyYear = Surveyyear)

Energy78 <- food78 %>% 
  select(ID, Country, SurveyYear, EnergykJ, Carbohydrateg, CHOpctotE)

@@

names(CW3CB3) <- c("Breakfast", "Morning.snack", "Lunch",
                   "Afternoon.snack", "Dinner", "Before.bedtime.snack",
                   "Midnight.food", "ID_DAY", "AGE", "SEX", "CPROB1",
                      "CPROB2", "CPROB3", "CPROB4", "CPROB5", "CPROB6", "CPROB7", "CPROB8", "CPROB9",
                      "CB", "CW", "MLCJOINT", "ID")
CW3CB3_reg <- CW3CB3[!duplicated(CW3CB3$ID), ]

>>>>

Energy14 <- food14 %>% 
  select(ID, Country, SurveyYear, EnergykJ, Carbohydrateg, CHOpctotE, Proteing, ProteinpctotE, 
         Fatg, FatpctotE)
Energy14$Country[Energy14$Country == "Northern Ireland"] <- "NI"

Energy56 <- food56 %>% 
  select(ID, Country, Surveyyear, EnergykJ, Carbohydrateg, CHOpctotE, Proteing, ProteinpctotE, 
         Fatg, FatpctotE) %>% 
  rename(SurveyYear = Surveyyear)

Energy78 <- food78 %>% 
  select(ID, Country, SurveyYear, EnergykJ, Carbohydrateg, CHOpctotE, Proteing, ProteinpctotE, 
         Fatg, FatpctotE)

@@

names(CW3CB3) <- c("Breakfast", "Morning.snack", "Lunch",
                   "Afternoon.snack", "Dinner", "Before.bedtime.snack",
                   "Midnight.food", "ID_DAY", "AGE", "SEX", "CPROB1",
                      "CPROB2", "CPROB3", "CPROB4", "CPROB5", "CPROB6", "CPROB7", "CPROB8", "CPROB9",
                      "CB", "CW", "MLCJOINT", "ID")

dta_NDNS <- read_csv("/home/wangcc-me/Documents/LSHTMproject/Rcode/dta_NDNS.csv") # extract the day of week data

CW3idday <- CW3CB3 %>% 
  select(ID, ID_DAY, CW, CB, MLCJOINT, AGE, SEX)

CW3idday$DayNo <- ave(CW3idday$ID_DAY, CW3idday$ID, FUN = seq_along) # adding the day no

dta_DayofWeek <- dta_NDNS %>% 
  select(id, id_dy, DayofWeek)

dta_DayofWeek$DayNo <- unlist(strsplit(dta_DayofWeek$id_dy,"D"))[c(FALSE, TRUE)] # creating a day no
names(dta_DayofWeek)[1] <- "ID"
dta_DayofWeek$DayNo <- as.numeric(dta_DayofWeek$DayNo)

CW3idday <-   CW3idday %>%  
  left_join(dta_DayofWeek, by= c("ID", "DayNo"))

CW3idday$DayofWeek[CW3idday$ID_DAY == 40714261000] <- "Saturday"
CW3idday$DayofWeek[CW3idday$ID_DAY == 112050710000] <- "Tuesday"
CW3idday$DayofWeek[CW3idday$ID_DAY == 310122510000] <- "Wednesday"
CW3idday$DayofWeek[CW3idday$ID_DAY == 5050616100] <- "Monday"
CW3idday$DayofWeek[CW3idday$ID_DAY == 50506161000] <- "Tuesday"
CW3idday$DayofWeek[CW3idday$ID_DAY == 505061610000] <- "Wednesday"
CW3idday$DayofWeek[CW3idday$ID_DAY == 7090824100] <- "Monday"
CW3idday$DayofWeek[CW3idday$ID_DAY == 70908241000] <- "Tuesday"
CW3idday$DayofWeek[CW3idday$ID_DAY == 709082410000] <- "Wednesday"
CW3idday$DayofWeek[CW3idday$ID_DAY == 807021910000] <- "Monday"
CW3idday$DayofWeek[CW3idday$ID_DAY == 812101310000] <- "Sunday"

with(CW3idday, tabpct(DayofWeek, CW))
with(CW3idday, chisq.test(DayofWeek, CW))

CW3idday$Weekend <- ifelse((CW3idday$DayofWeek == "Sunday" | CW3idday$DayofWeek == "Saturday"), TRUE, FALSE)

with(CW3idday, tabpct(Weekend, CW))
with(CW3idday, chisq.test(Weekend, CW))

CW3CB3_reg <- CW3CB3[!duplicated(CW3CB3$ID), ]
